\section*{Abstract}
\label{sec:abstract}

Understanding how molecular programs are embedded within tissue morphology is a central challenge in spatial biology. While vision transformer (ViT) foundation models capture rich histological structure and spatial transcriptomics provide molecular context, existing multimodal approaches largely rely on contrastive alignment and do not directly learn joint morpho-molecular representations. We introduce a family of multimodal transformer architectures that integrate subcellular Xenium transcript readouts with H\&E histology at varying levels of fusion granularity, spanning \emph{early-fusion} (transcript tokens injected directly into the ViT stream) and \emph{late-fusion} variants (embedding-level combination of independently encoded modalities), each available with element-wise addition (\texttt{add}) or concatenation (\texttt{concat}) operators. Token-level transcript encoding is compared against tile-level aggregation, yielding six multimodal configurations evaluated alongside unimodal baselines (\textit{vision}, \textit{expr-tile}, \textit{expr-token}, CONCH, Geneformer).

We benchmark all models on two complementary supervised tasks: (i)~\emph{gene expression prediction}, where held-out genes are predicted from a partial transcript panel and histology; and (ii)~\emph{cell type composition prediction}, a higher-order task requiring capture of local co-expression patterns. Experiments are conducted on three datasets---our internal BEAT cancer cohort, the public HEST1k compendium, and the HESCAPE benchmark---across two gene panels (50 most highly variable genes; HESCAPE panel) and two model complexities (low: ViT-S with a 1-layer MLP; high: ViT-B with a 3-layer MLP). We further evaluate downstream clinical utility via multiple-instance learning (MIL) on histology classification and survival prediction using embeddings from all model variants.

Across all settings, multimodal fusion consistently outperforms unimodal baselines, with the largest gains on cell type prediction. Token-level transcript representations are a critical driver of performance, outperforming tile-level aggregation independent of fusion timing. Model capacity selectively benefits multimodal models, implicating cross-modal interaction as the primary driver of additional representational capacity. Spatial predictions faithfully recapitulate the ground-truth distribution of gene expression and cellular composition across tissue sections, and MIL experiments demonstrate that richer morpho-molecular representations translate to improved clinical endpoint prediction.

